---
title: "Stat Project"
subtitle: ""
authors: "Ema, Eileen, Elsie"
format: pdf
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    fig.width: 4
    fig.height: 3
---

Initial setup:
```{r}
library(tidyverse)
library(ggplot2)

proj_data <- read.csv("sample_101_Life-Expectancy-Data-Updated.csv") |>
  mutate(Economy_status_Developed = as.factor(Economy_status_Developed))
```

## Introduction
1. General
Life expectancy is primarily regarded as the main indicator for views on the general health of certain populations. A major part of evaluating life expectancy is knowing which other variables could have an impact on the life expectancy, and how changes in one variable affects life expectancy. The primary goal of this project is to evaluate which demographic / social health-related factors are associated with variation in life expectancy across different countries and years. Specifically, there are many individual factors such as years in schooling, percentage of measles, and economy status. Previous research, including a study published in 2018 by Daniel et. al, have emphasized the role addressing social determinants can improve patient care. Daniel et. al claimed that there is a "15-year difference in life expectancy between the most advantaged and disadvantaged citizens". Based on this information and the dataset we chose, we wanted to explore along a similar line if there are specific factors that could affect life expectancy more than other factors.

2. Dataset
The dataset used in this project came originally from Kaggle, which was sourced from the WHO (World Health Organization). The actual data was compiled from across three sources: WHO, World Bank data, and Our World in Data - a University of Oxford project. Each row represents a country and all the data collected from one year in that country across all the other variables. Even though the dataset comes from multiple sources, they are all adjusted and standardized. The observational unit was a country-year. All the dataset variables are listed below:

Dataset variables:

Country: A list of 179 distinct countries included in the dataset.
Region: Categorizes the 179 countries into 9 geographical regions, such as Africa, Asia, Oceania, and the European Union.
Year: The observed year, ranging from 2000 to 2015.
Infant_deaths: Represents the number of infant deaths per 1,000 population.
Under_five_deaths: Represents the number of deaths of children under five years old per 1,000 population.
Adult_mortality: Represents the number of deaths of adults per 1,000 population.
Alcohol_consumption: Records alcohol consumption in litres of pure alcohol per capita for individuals aged 15 years and over.
Hepatitis_B: Represents the percentage of coverage for Hepatitis B (HepB3) immunisation among 1-year-olds.
Measles: Represents the percentage of coverage for Measles containing vaccine first dose (MCV1) immunisation among 1-year-olds.
BMI: Body Mass Index, a measure of nutritional status in adults (defined as a person's weight in kilograms divided by the square of that person's height in meters).
Polio: Represents the percentage of coverage for Polio (Pol3) immunisation among 1-year-olds.
Diphtheria: Represents the percentage of coverage for Diphtheria tetanus toxoid and pertussis (DTP3) immunisation among 1-year-olds.
Incidents_HIV: Represents the incidents of HIV per 1,000 population aged 15-49.


Response Variable:
Life_expectancy: The average life expectancy for both genders across different years, from 2000 to 2015.

Predictor Variables:
Economic and Demographic --

GDP_per_capita: Gross Domestic Product per capita in current US Dollars.
Population_mln: Total population of a country in millions.
Schooling: Average years individuals aged 25 and over have spent in formal education.
Economy_status_Developed: A binary indicator (0 or 1) denoting whether a country is classified as 'Developed'.
Economy_status_Developing: A binary indicator (0 or 1) denoting whether a country is classified as 'Developing'.

Lifestyle --
Alcohol_consumption: Records alcohol consumption in litres of pure alcohol per capita for individuals aged 15 years and over.
BMI: Body Mass Index, a measure of nutritional status in adults (defined as a person's weight in kilograms divided by the square of that person's height in meters).
Thinness_ten_nineteen_years: Prevalence of thinness among adolescents aged 10-19 years (specifically, BMI < -2 standard deviations below the median).
Thinness_five_nine_years: Prevalence of thinness among children aged 5-9 years (specifically, BMI < -2 standard deviations below the median).

Mortality and Disease --
Infant_deaths: Represents the number of infant deaths per 1,000 population.
Under_five_deaths: Represents the number of deaths of children under five years old per 1,000 population.
Adult_mortality: Represents the number of deaths of adults per 1,000 population.
Hepatitis_B: Represents the percentage of coverage for Hepatitis B (HepB3) immunisation among 1-year-olds.
Measles: Represents the percentage of coverage for Measles containing vaccine first dose (MCV1) immunisation among 1-year-olds.
Polio: Represents the percentage of coverage for Polio (Pol3) immunisation among 1-year-olds.
Diphtheria: Represents the percentage of coverage for Diphtheria tetanus toxoid and pertussis (DTP3) immunisation among 1-year-olds.
Incidents_HIV: Represents the incidents of HIV per 1,000 population aged 15-49.

Observational Units:
Country: A list of 179 distinct countries included in the dataset.
Region: Categorises the 179 countries into 9 geographical regions, such as Africa, Asia, Oceania, and the European Union.
Year: The observed year, ranging from 2000 to 2015.

3. Rationale
The reason for this data analysis is that if helps identify which health and economic variables most strongly explain differences in life expectancy. It also helps shows how different diseases / vaccinations and child and adult mortality can contribute to the overall health of a nation. This data analysis can also help us make conclusions on what types of health policy can be recommended or suggested in order to target preventable health risks in certain countries.

4. Exploratory Data Analysis
```{r}
library(tidyverse)
library(ggplot2)

ggplot(data = proj_data, mapping = aes(x = Life_expectancy)) + 
  geom_histogram() + labs(title = "Distribution of Life Expectancy", x = "Life Expectancy (years)", y = "Frequency")
```

```{r}
ggplot(data = proj_data, mapping = aes(x = GDP_per_capita, y = Life_expectancy)) +
  geom_point(alpha = 0.5) +
  labs(title = "GDP per Capita vs. Life Expectancy", x = "GDP per Capita", y = "Life Expectancy (years)")
```

```{r}
ggplot(data = proj_data, mapping = aes(x = Adult_mortality, y = Life_expectancy)) +
  geom_point(alpha = 0.5) +
  labs(title = "Adult Mortality Rate vs. Life Expectancy", x = "Adult Mortality Rate", y = "Life Expectancy (years)")
```
The EDA helps us visualize some of the aspects of this dataset. It helps look at life expectancy as a whole across all the countries, and to help see if it is normally distributed. Additionally, the two scatter plots visually confirm whether economic indicators (GDP per capita) and health indicators (adult mortality) are strong predictors of life expectancy.

## Methodolody
WORDS HERE


Assumptions:
```{r}
library(tidyverse)
library(ggplot2)
m1 <- lm(Life_expectancy ~ Year + Infant_deaths + Under_five_deaths + Adult_mortality + Alcohol_consumption + Hepatitis_B + Measles + BMI + Polio + Diphtheria + Incidents_HIV + GDP_per_capita + Population_mln + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed + Region, data = proj_data) 
ggplot(data = proj_data, mapping = aes(x = m1$fitted.values, y = m1$residuals)) + 
  geom_point() + 
  labs(x = "Fitted values", y = "Residuals",
       title = "Fitted Values vs Residuals") + 
  geom_hline(yintercept = 0, color = "red") # add a horizontal red line at 0

ggplot(data = proj_data, mapping = aes(x = m1$residuals)) + 
  geom_histogram(color = "blue",
                 fill = "blue") + 
  labs(x = "Residuals", y = "Frequency",
       title = "Histogram of Residuals")
```

## Results
```{r}
m1 <- lm(Life_expectancy ~ Year + Infant_deaths + Under_five_deaths + Adult_mortality + Alcohol_consumption + Hepatitis_B + Measles + BMI + Polio + Diphtheria + Incidents_HIV + GDP_per_capita + Population_mln + Thinness_ten_nineteen_years + Thinness_five_nine_years + Schooling + Economy_status_Developed + Region, data = proj_data) 

library(broom)
linear_regression <- tidy(m1) |>
  select(term, estimate, p.value) |>
  rename(
    Variable = term,
    Estimate = estimate,
    P_value = p.value) |>
  mutate(
    Estimate = round(Estimate, 3),
    P_value = round(P_value, 3))
print(linear_regression)

```

## Discussion

## Citations
https://www.acpjournals.org/doi/10.7326/M17-2441
https://www.opendatabay.com/data/ai-ml/0bdbea7e-f40b-4c41-b010-37537d03a723?utm_source